---
title: "Modelling loan interest rates"
format: html
editor: visual
---


In this application exercise we will be studying loan interest rates.
The dataset is one you've come across before in your reading -- the dataset about loans from the peer-to-peer lender, Lending Club, from the **openintro** packahe.
We will use **tidyverse** and **tidymodels** for data exploration and modeling, respectively.

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(tidymodels)
library(openintro)
```

Before we use the data set, we'll make a few transformations to it.

Warm up: What is this code doing?

```{r}
loans <- loans_full_schema %>%
  mutate(
    credit_util = total_credit_utilized / total_credit_limit,
    bankruptcy  = as.factor(if_else(public_record_bankrupt == 0, 0, 1)),
    verified_income = droplevels(verified_income),
    homeownership = str_to_title(homeownership),
    homeownership = fct_relevel(homeownership, "Rent", "Mortgage", "Own")
    ) %>%
  rename(credit_checks = inquiries_last_12m) %>%
  select(interest_rate, verified_income, debt_to_income, credit_util, bankruptcy, term, credit_checks, issue_month, homeownership) 
```

Take a glimpse of the data below:

```{r}
#| label: glimpse-loans

```

# Interest rate vs. credit utilization ratio


Write a model that models interest rate by credit utilization ratio. Name it `rate_util_fit`.

```{r}
#| label: fit-rate-util

rate_util_fit <-
```

And here is the model visualized:

```{r}
ggplot(loans, aes(x = credit_util, y = interest_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")
```

-   **Your turn:** What is the estimated interest rate for a loan applicant with credit utilization of 0.8, i.e. someone whose total credit balance is 80% of their total available credit?

```{r}
#| label: predict-80


```

# Leverage vs Influence 

What is leverage? What is influence? 


Now, let's add data points to drive this idea home!


```{r}

new_point_1 <- data.frame(100, NA, NA, 0.8, NA,  NA, NA, NA, NA)
names(new_point_1) <- names(loans)

new_point_2 <- data.frame(100, NA, NA, 4, NA,  NA, NA, NA, NA)
names(new_point_2) <- names(loans)


new_loans1 <- rbind(loans, new_point_1)

new_loans2 <- rbind(loans, new_point_2)
```


Below, compare the model outputs. 


```{r}

linear_reg() |>
  set_engine("lm") |>
  fit(interest_rate ~ credit_util, data = new_loans) |>
  tidy()

linear_reg() |>
  set_engine("lm") |>
  fit(interest_rate ~ credit_util, data = new_loans1) |>
  tidy()

linear_reg() |>
  set_engine("lm") |>
  fit(interest_rate ~ credit_util, data = new_loans2) |>
  tidy()


```

```{r}
ggplot(loans, aes(x = credit_util, y = interest_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")
```


*Add Response*


# Interest rate vs. homeownership

Next lets predict interest rates from homeownership, which is a categorical predictor with three levels:

```{r}
levels(loans$homeownership)
```

-   **Demo:** Fit the linear regression model to predict interest rate from homeownership and display a tidy summary of the model. Write the estimated model output below.

```{r}
#| label: fit-rate-home

rate_util_home_fit <- 
  
```

-   **Your turn:** Interpret each coefficient in context of the problem.

    -   Intercept: 

    -   Slopes:

        -   

        -   

# Interest rate vs. credit utilization and homeownership

## Main effects model

-   **Demo:** Fit a model to predict interest rate from credit utilization and homeownership, **without** an interaction effect between the two predictors. Display the summary output and write out the estimated regression equation.

```{r}
#| label: rate-util-home-fit
rate_util_home_fit <- linear_reg() |>
  fit(interest_rate ~ credit_util + homeownership, data = loans)
tidy(rate_util_home_fit)
```

$$


$$

-   **Demo:** Write the estimated regression equation for loan applications from each of the homeownership groups separately.
    -   Rent: $ $
    -   Mortgage: $  $
    -   Own: $  $

- **Question:** How does the model predict the interest rate to vary as credit utilization varies for loan applicants with different homeownership status. Are the rates the same or different?


## Interaction effects model

- **Demo:** Fit a model to predict interest rate from credit utilization and homeownership, **with** an interaction effect between the two predictors. Display the summary output and write out the estimated regression equation.

```{r}
#| label: rate-util-home-fit
rate_util_home_int_fit <- 
```

$$
\widehat{interest~rate} = 9.44 + 6.20 \times credit\_util + 1.39 \times Mortgage + 0.697 \times Own - 1.64 \times credit\_util:Mortgage - 1.06 \times credit\_util:Own
$$

-   **Demo:** Write the estimated regression equation for loan applications from each of the homeownership groups separately.
    -   Rent: $\widehat{interest~rate} = 9.44 + 6.20 \times credit~util$
    -   Mortgage: $\widehat{interest~rate} = 10.83 + 4.56 \times credit~util$
    -   Own: $\widehat{interest~rate} = 10.137 + 5.14 \times credit~util$

- **Question:** How does the model predict the interest rate to vary as credit utilization varies for loan applicants with different homeownership status. Are the rates the same or different?


## Choosing a model

Rule of thumb: **Occam's Razor** - Don't overcomplicate the situation!
We prefer the *simplest* best model.

```{r}
#| label: glance
glance(rate_util_home_fit)
glance(rate_util_home_int_fit)
```

-   **Review:** What is R-squared? What is adjusted R-squared?

R-squared is 

Adjusted R-squared is 

- **Question:** Based on the $R^2s$ of these two models, which one do we prefer? Is this appropriate?

- **Question:** Based on the adjusted $R^2$s of these two models, which one do we prefer?




Note: We can use the principles of adjusted R-squared to select models using forwards or backwards selection (Read 8.4.1)

# Another model to consider

- **Your turn:** Let's add one more model to the variable -- issue month. Should we add this variable to the interaction effects model from earlier?

```{r}
linear_reg() |>
  fit(interest_rate ~ credit_util * homeownership + issue_month, data = loans) |>
  glance()
```


# Optional 

## 2 Quantitative Explanatory Variables

Remember the penguins? Let's finish the activity. 

Now, let's explore body mass, and it's relationship to bill length and flipper length.

```{r}
#| message: false
#| warning: false

library(palmerpenguins)
library(plotly)
library(widgetframe)
```


Note: This code is beyond the scope of this course!

```{r}

quanplot <- plot_ly(penguins, 
                    x = ~ flipper_length_mm, y = ~ bill_length_mm, z = ~body_mass_g,
                    marker = list(size = 3, color = "lightgray" , alpha = 0.5, 
                                  line = list(color = "gray" , width = 2))) |>
                      add_markers() |>
                      plotly::layout(scene = list(
                        xaxis = list(title = "Flipper (mm)"),
                        yaxis = list(title = "Bill (mm)"), 
                        zaxis = list(title = "Body Mass (g)")
                      )) |>
                    config(displayModeBar = FALSE)
                  frameWidget(quanplot)
                      

```

-- Fit the additive model below. Interpret the coefficient for flipper in context of the problem.

```{r}



```



What if I want to fit an interaction model with these two quantitative variables?

It looks really similar to what we've done before! Try it out!